---
title: "Lab 5"
author: "Antonio D'Alessandro"
output: pdf_document
date: 
---


We will work with the diamonds dataset from last lecture:

```{r}
pacman::p_load(ggplot2) #this loads the diamonds data set too
?diamonds
diamonds$cut =      factor(diamonds$cut, ordered = FALSE)
diamonds$color =    factor(diamonds$color, ordered = FALSE)
diamonds$clarity =  factor(diamonds$clarity, ordered = FALSE)
skimr::skim(diamonds)



```

Given the information above, what are the number of columns in the raw X matrix?

9

Verify this using code:

```{r}

ncol(diamonds)-1

```

Would it make sense to use polynomial expansions for the variables cut, color and clarity? Why or why not?

No, they are nominal variables. If we dummified the nomial/categorical variables it would not make sense.

Would it make sense to use log transformations for the variables cut, color and clarity? Why or why not?



In order to ensure there is no time trend in the data, randomize the order of the diamond observations in D:.

```{r}

diamonds <- diamonds[sample(1:nrow(diamonds)),]

```

Let's also concentrate only on diamonds with <= 2 carats to avoid the issue we saw with the maximum. So subset the dataset. Create a variable n equal to the number of remaining rows as this will be useful for later. Then plot it.

```{r}

diamonds <- diamonds[diamonds$carat<=2,]

n= nrow(diamonds)
ggplot(diamonds, aes(x = carat, y = price)) + 
  geom_point()
```

Create a linear model of price ~ carat and gauge its in-sample performance using s_e.

```{r}

mod <- lm(price ~ carat, diamonds)
summary(mod)$sigma
```

Create a model of price ~ clarity and gauge its in-sample performance

```{r}

mod2 <- lm(price ~ clarity, diamonds)
summary(mod2)$sigma

```

Why is the model price ~ carat substantially more accurate than price ~ clarity?

Because clarity is a categorical variable a thus more difficult to get an accurate handle on what is happening with vs. a continuous predictor. 

Create a new transformed feature ln_carat and plot it vs price.

```{r}

diamonds$ln_carat <- log(diamonds$carat)

ggplot(diamonds, aes(x = ln_carat, y = price)) + 
  geom_point()

ggplot(diamonds, aes(x = carat, y = price)) + 
  geom_point()


```

Would price ~ ln_carat be a better fitting model than price ~ carat? Why or why not?

No it will not be better, a linear relationship will capture better you can see if you compare two plots like in lines 82-92. 

Verify this by comparing R^2 and RMSE of the two models:

```{r}
mod1 <- lm(price ~ carat, diamonds)
summary(mod1)$sigma

mod2 <- lm(price ~ ln_carat, diamonds)
summary(mod2)$sigma
```

Create a new transformed feature ln_price and plot its estimated density:


```{r}
diamonds$ln_price <- log(diamonds$price)


ggplot(diamonds) + geom_histogram(aes(x = ln_price), binwidth = 0.01)
```


Now plot it vs carat.

```{r}
ggplot(diamonds, aes(x = carat, y = ln_price)) + 
  geom_point()
```

Would ln_price ~ carat be a better fitting model than price ~ carat? Why or why not?

Going to guess ln_price ~ carat will be better by comparing the two ggplots

Verify this by computing s_e of this new model. Make sure these metrics can be compared apples-to-apples with the previous.

```{r}

mod3 <- lm(ln_price ~ carat, diamonds)
yhat <- exp(mod3$fitted.values)
SSE <- sum((yhat-diamonds$price)^2)
RMSE <- sqrt(SSE/(n-2))
RMSE
summary(mod1)$sigma

```

We just compared in-sample statistics to draw a conclusion on which model has better performance. But in-sample statistics can lie! Why is what we did likely valid?

We are only using one feature with n greater than 50,000 no possible way of overfitting. 

Plot ln_price vs ln_carat.

```{r}
ggplot(diamonds, aes(x = ln_carat, y = ln_price)) + 
  geom_point()
```

Would ln_price ~ ln_carat be the best fitting model than the previous three we considered? Why or why not?

Yes because it looks the most linear. 

Verify this by computing s_e of this new model. Make sure these metrics can be compared apples-to-apples with the previous.

```{r}


mod4 <- lm(ln_price ~ ln_carat, diamonds)
yhat <- exp(mod4$fitted.values)
SSE <- sum((yhat-diamonds$price)^2)
RMSE <- sqrt(SSE/(n-2))
RMSE
summary(mod4)$sigma



```

Compute b, the OLS slope coefficients for this new model of ln_price ~ ln_carat.

```{r}
coef(mod4)

```

Interpret b_1, the estimated slope of ln_carat.

Percent change in price is equal to approx b_1 * percent change in carat

Interpret b_0, the estimated intercept.

Extrapolating here, no meaning. Predicted log(price) of a diamond with 0 weight.

Create other features ln_x, ln_y, ln_z, ln_depth, ln_table.

```{r}

diamonds$ln_x <- log(diamonds$x)
diamonds$ln_y <- log(diamonds$y)
diamonds$ln_z <- log(diamonds$z)
diamonds$ln_depth <- log(diamonds$depth)
diamonds$ln_table <- log(diamonds$table)




```

From now on, we will be modeling ln_price (not raw price) as the prediction target. 

Create a model (B) of ln_price on ln_carat interacted with clarity and compare its performance with the model (A) ln_price ~ ln_carat.

```{r}


modb<- lm(ln_price ~ ln_carat * clarity, diamonds)
moda <- lm(ln_price ~ ln_carat, diamonds)

summary(modb)$sigma
summary(moda)$sigma
```

Which model does better? Why?

Model B did better, adding the interaction with clarity added complexity to H with got us closer to f. 

Create a model of (C) ln_price on ln_carat interacted with every categorical feature (clarity, cut and color) and compare its performance with model (B)

```{r}

modc <- lm(ln_price ~ ln_carat * (clarity + cut + color), diamonds)

summary(modc)$sigma

```

Which model does better? Why?

Model C did better than both A and B, for the same reason as why B did better than A. We are adding more useful dimensions to the colsp of X / increasing complexity to H which is getting us closer to f. 

Create a model (D) of ln_price on every continuous feature (logs of carat, x, y, z, depth, table) interacted with every categorical feature (clarity, cut and color) and compare its performance with model (C).

```{r}

diamonds <- diamonds[diamonds$x >0 & diamonds$z >0,]
modd <- lm(ln_price ~ (ln_carat + ln_x + ln_y + ln_z + ln_depth + ln_table) * (clarity + cut + color), diamonds)

summary(modd)$sigma
```

Which model does better? Why?

Model D did better than all the others for the same reason as the previous questions. We are adding more useful dimensions (non-orthogonal) to the colspace of X / increasing complexity to H which is getting us closer to f. 


What is the p of this model D? Compute with code.

```{r}

modd$rank

ncol(model.matrix(~(ln_carat + ln_x + ln_y + ln_z + ln_depth + ln_table) * (clarity + cut + color), diamonds))

```

Create model (E) which is the same as before except create include the raw features interacted with the categorical features and gauge the performance against (D).

```{r}

mod_e <- lm(ln_price ~ (carat + x + y + z + depth + table) * (clarity + cut + color), diamonds)
summary(mod_e)$sigma

```

Which model does better? Why?

The non-linearities do not impact as much as the raw. Linear slopes are better than log slopes. 

Create model (F) which is the same as before except also include also third degree polynomials of the continuous features interacted with the categorical features and gauge performance against (E). By this time you're getting good with R's formula syntax!

```{r}

modf <- lm(ln_price ~ (poly(carat,3) + poly(x,3) + poly(y,3) + poly(z,3) + poly(depth,3) + poly(table,3)) * (clarity + cut + color), diamonds)
summary(modf)$sigma
```

Which model does better? Why?

F does better.

Can you think of any other way to expand the candidate set curlyH? Discuss.



We should probably assess oos performance now. Sample 2,000 diamonds and use these to create a training set of 1,800 random diamonds and a test set of 200 random diamonds. Define K and do this splitting:

```{r}

K <- 10 
set.seed(1984)
nsamp <- 2000
D <- diamonds[sample(1:nrow(diamonds),nsamp),]
D_train <- D[1:((1-1/K)*nsamp),]
D_test <- D[((1-1/K)*nsamp+1):nsamp,]


```

Compute in and out of sample performance for models A-F. Use s_e as the metric (standard error of the residuals). Create a list with keys A, B, ..., F to store these metrics. Remember the performances here will be worse than before since before you're using nearly 52,000 diamonds to build a model and now it's only 1,800! 

```{r}


insampleRMSE <- list()
oosRMSE <- list()


moda <- lm(ln_price ~ ln_carat, D_train)
insampleRMSE[['A']] <- summary(moda)$sigma

modb<- lm(ln_price ~ ln_carat * clarity, D_train)
insampleRMSE[['B']] <-summary(modb)$sigma

modc <- lm(ln_price ~ ln_carat * (clarity + cut + color), D_train)
insampleRMSE[['C']] <-summary(modc)$sigma

modd <- lm(ln_price ~ (ln_carat + ln_x + ln_y + ln_z + ln_depth + ln_table) * (clarity + cut + color), D_train)
insampleRMSE[['D']] <-summary(modd)$sigma

mod_e <- lm(ln_price ~ (carat + x + y + z + depth + table) * (clarity + cut + color), D_train)
insampleRMSE[['E']] <-summary(mod_e)$sigma

modf <- lm(ln_price ~ (poly(carat,3) + poly(x,3) + poly(y,3) + poly(z,3) + poly(depth,3) + poly(table,3)) * (clarity + cut + color), D_train)
insampleRMSE[['F']] <-summary(modf)$sigma


oosRMSE[['A']] <- sd(D_test$ln_price - predict(moda,D_test))

oosRMSE[['B']] <- sd(D_test$ln_price - predict(modb,D_test))

oosRMSE[['C']] <- sd(D_test$ln_price - predict(modc,D_test))

oosRMSE[['D']] <- sd(D_test$ln_price - predict(modd,D_test))

oosRMSE[['E']] <- sd(D_test$ln_price - predict(mod_e,D_test))

oosRMSE[['F']] <- sd(D_test$ln_price - predict(modf,D_test))


cbind(
  unlist(insampleRMSE),
  unlist(oosRMSE)
)


```

You computed oos metrics only on n_* = 200 diamonds. What problem(s) do you expect in these oos metrics?

Bias / Variance trade off. Since K was only 10 (10% of data going to test) we will have greater bias but less variance. We will be further away from the true oosRMSE but more sure about what are value is (less variance). 

To do the K-fold cross validation we need to get the splits right and crossing is hard. I've developed code for this already. Run this code.

```{r}
temp = rnorm(n)
folds_vec = cut(temp, breaks = quantile(temp, seq(0, 1, length.out = K + 1)), include.lowest = TRUE, labels = FALSE)
head(folds_vec, 200)
```

Comment on what it does and how to use it to do a K-fold CV:

This chunk of code creates the various K values to use for K-fold validation and stores them in a vector. <- not what it is 

Do the K-fold cross validation for model F and compute the overall s_e and s_s_e. 

```{r}

set.seed(1986)
nsamp <- 2000
D <- diamonds[sample(1:nrow(diamonds),nsamp),]
s_e <- c()



#create folds stored in matrix  
kfolds <- matrix(1:2000, ncol=10)
#create empty vector to store individual s_e's
vals <- c()

#iterate through matrix columns for k-fold validation
for(i in 1:ncol(kfolds))  {

D_train <- subset(D, subset = row.names(D) != mx[,i])  
D_test <- subset(D, subset = row.names(D) == mx[,i])

modf <- lm(ln_price ~ (poly(carat,3) + poly(x,3) + poly(y,3) + poly(z,3) + poly(depth,3) + poly(table,3)) * (clarity + cut + color), D_train)
s_e <- c(sd(D_test$ln_price - predict(modf,D_test)))

vals <- append(vals,s_e)

}

s_e <- mean(vals)

s_s_e <- sd(vals)


```

Does K-fold CV help reduce variance in the oos s_e? Discuss.

Yes K-fold reduces variance in our oos error metrics. We are averaging many realizations of the oos metric which is a random variable (it is an r.v. because the selection of the test/train data split is randomized) 

Imagine using the entire rest of the dataset besides the 2,000 training observations divvied up into slices of 200. Measure the oos error for each slice on Model F in a vector s_e_s_F and compute the s_s_e_F and also plot it.

```{r}



ggplot(data.frame(s_e_s_F = s_e_s_F)) + geom_histogram(aes(x = s_e_s_F))
```